# Eigenfaces

```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, messgae = FALSE)
```

So what is an [Eigenface](https://en.wikipedia.org/wiki/Eigenface)? Let's fine out!

We follow [this blogpost](https://towardsdatascience.com/eigenfaces-recovering-humans-from-ghosts-17606c328184) and recreate the approach in `R`.

First off let's get a sample of the data from the `statbiscuits` repository on GitHub (these data are wrangled versions of those from [`n0acar`'s GitHub repository](https://github.com/n0acar/eigenfaces/tree/master/lfw_cropped_gray_eigenfaces)), see [here](https://github.com/cmjt/statbiscuits/blob/master/swots/raw-scripts/raw_eigenfaces.r for the raw script I used to download and wrangle the data.

```{r get data from GH, eval = FALSE}
data_url <- "https://github.com/cmjt/statbiscuits/raw/master/swots/data/faces.RData"
load(url(data_url))
```

```{r sneak face data, echo = FALSE}
load("data/faces.RData")
```

The `faces` object is a `list()` with `r length(faces)` named elements:

```{r data strucure}
library(pixmap) ## needed to deal with the data
names(faces)
```

Each element is a grey pixmap image of the person named. Let's plot the pixmap images

```{r plotfaces, fig.width = 8,fig.height = 8, fig.cap = "Each face in our dataset.", fig.align = "center" }
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(faces)){
    plot(faces[[i]])
    legend("bottom", bty = "n", names(faces[i]))
}
```

## Prep for PCA

Collapse each image into one column of a matrix.

```{r collapse}
## collapse the grey matrix
face_data <- lapply(faces, function(x) c(x@grey))
## each column is a face
face_data <- do.call('cbind',face_data)
colnames(face_data) <- names(faces)
## have a look at the data matrix
head(face_data)
```

### Center around mean

```{r meanface, fig.cap = "The mean face from the 16 in our dataset.", fig.align = "center" ,fig.width = 8, fig.height = 8}
## center around mean
mean <- apply(face_data, 1, mean)
mean_face <- pixmapGrey(mean, 64, 64, bbox = c(0, 0, 64, 64))
plot(mean_face)
```

```{r center, fig.width = 8,fig.height = 8, fig.cap = "Each face centered (i.e., with the mean face subtracted).", fig.align = "center" }
centered <- apply(face_data, 2, function(x) x - mean)
## centered images
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(faces)){
    plot(pixmapGrey(centered[,i], 64, 64, bbox = c(0, 0, 64, 64)))
    legend("bottom", bty = "n", paste(names(faces[i]), "centered", sep = " "))
}
```

## Principal Component Analysis (PCA)

```{r pca}
pca <- prcomp(centered)
summary(pca)

```


The PCA object, `pca`, contains the following information:

  - the center point (`$center`), 
  - scaling (`$scale`), and
  - standard deviation (`$sdev`) of each principal component, as well as
   - the principal components (`$rotation`), and
   - the values of each sample in terms of the principal components (`$x`)



```{r biplot, fig.height=8,fig.width=8, fig.align="center"}
## for a nice biplot
library(ggfortify)
library(ggplot2)
## using autoplot
## PC1 vs PC2 by default
autoplot(pca,loadings = TRUE,loadings.label = TRUE,alpha = 0.1)
## play around with the arguments to see what they control
autoplot(pca,x = 3, y = 4,loadings = TRUE,loadings.label = TRUE,alpha = 0.1)
```

### Eigenfaces (Principal Components)

```{r efaces,  fig.width = 8,fig.height = 8, fig.cap = "Each eigenface (PC1 to PC16). Note how noisy the faces get."}
## eigen faces
pcs <- pca$x
eigenfaces <- apply(pcs,2, function(x)  pixmapGrey(x, 64, 64, bbox = c(0, 0, 64, 64)))
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(eigenfaces)){
    plot(eigenfaces[[i]])
    legend("bottom", bty = "n", paste("PC", i, sep = " "))
}
```

### Reconstructing

#### How many coponents should we keep?
 
Each face is a weighted (loadings) combintation of the eigen faces (PCs) above, but how many PCs (eigenfaces) do we keep?
 
```{r screeplot, fig.align="center"}
## screeplot
screeplot(pca,type = "lines", pch = 20)
```


```{r loadings}
str(pca$rotation)
```

```{r recon, fig.width = 8,fig.height = 8}
## 2 PCs
n_pc <- 2

recon <- pixm <- list()
for (i in 1:length(faces)){
    recon[[i]] <- rowSums(pca$x[,1:n_pc]%*%pca$rotation[i,1:n_pc])
    pixm[[i]] <- pixmapGrey(recon[[i]], 64, 64, bbox = c(0, 0, 64, 64))
}

## 2PC to reconstruct centered imags
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(pixm)){
    plot(pixm[[i]])
    legend("bottom", bty = "n", paste(names(faces[i]), "2 PC", sep = " "))
}

### Hmm OK, what about 5
## 5 PCs
n_pc <- 5

recon <- pixm <- list()
for (i in 1:length(faces)){
    recon[[i]] <- rowSums(pca$x[,1:n_pc]%*%pca$rotation[i,1:n_pc])
    pixm[[i]] <- pixmapGrey(recon[[i]], 64, 64, bbox = c(0, 0, 64, 64))
}

## 5PC to reconstruct centered images
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(pixm)){
    plot(pixm[[i]])
    legend("bottom", bty = "n", paste(names(faces[i]), "5 PC", sep = " "))
}

```

```{r recon_full,  fig.width = 8,fig.height = 8}
### Hmm OK, what about 10
## 10 PCs
n_pc <- 10

recon <- pixm <- list()
for (i in 1:length(faces)){
    recon[[i]] <- rowSums(pca$x[,1:n_pc]%*%pca$rotation[i,1:n_pc])
    pixm[[i]] <- pixmapGrey(recon[[i]], 64, 64, bbox = c(0, 0, 64, 64))
}
## 10PC to reconstruct centered images
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(pixm)){
    plot(pixm[[i]])
    legend("bottom", bty = "n", paste(names(faces[i]), "10 PC", sep = " "))
}
## remember this is the centered image, let's add the "mean face" to reconstruct the original
par(mfrow = c(4,4), mar = c(0,0,0,0), oma = c(0,0,0,0))
for(i in 1:length(recon)){
    plot(pixmapGrey(recon[[i]] + mean, 64, 64, bbox = c(0, 0, 64, 64)))
    legend("bottom", bty = "n", paste(names(faces[i]), "10 PC + mean", sep = " "))
}
```


## 1000 faces

Above we use only 16 images, but what if we had many, many, more. Then dimension reduction becomes much more useful. [Here is a zipped archive](http://conradsanderson.id.au/lfwcrop/lfwcrop_grey.zip) of over 13,000 faces as grey pixmap images, as above.

We can download the archive, unzip, and read the files into `R`:

```{r zipped}
zip <- "http://conradsanderson.id.au/lfwcrop/lfwcrop_grey.zip"
temp <- tempfile(fileext = ".zip")
download.file(zip,destfile = temp)
tst <- unzip(temp)
unlink(temp)
all_faces <- list.files("lfwcrop_grey/faces",pattern = ".pgm", full = TRUE)
## read in a random subset of these images
## using set.seed to ensure the following is reproducible (means you will get the same subset)
set.seed(6666)
choose <- sample(1:length(all_faces), 1000, replace = FALSE)
face_set <- lapply(choose, function(x) read.pnm(all_faces[x]))
## name the object accordingly
col <- sapply(all_faces[choose], function(x) strsplit(strsplit(x,"faces/")[[1]][2],"_")[[1]])
names <-  sapply(col, function(y) paste(y[-length(y)],  collapse = " ")); names(names) <- NULL
head(names) ## if you've used the same seed as above this is what you should see
names(face_set) <- names
```

Mean face (of the 1000 faces)

```{r collapse and meanface, fig.cap = "The mean face from the 1000 in our dataset. SPOOKY", fig.align = "center", fig.width = 8, fig.height = 8}
## collapse the grey matrix
face_data <- lapply(face_set, function(x) c(x@grey))
## each column is a face
face_data <- do.call('cbind',face_data)
colnames(face_data) <- names(face_set)
## mean face
mean <- apply(face_data, 1, mean)
mean_face <- pixmapGrey(mean, 64, 64, bbox = c(0, 0, 64, 64))
plot(mean_face)
```

PCA

```{r pca all}
centered <- apply(face_data, 2, function(x) x - mean)
pca <- prcomp(centered)
## screeplot
screeplot(pca,type = "lines", pch = 20)
```

What does our first eigen face look like?

```{r eface,  fig.width = 8,fig.height = 8, fig.cap = "First eigenface (PC1)."}
## eigen faces
pcs <- pca$x
eigenfaces <- apply(pcs,2, function(x)  pixmapGrey(x, 64, 64, bbox = c(0, 0, 64, 64)))
plot(eigenfaces[[1]])
legend("bottom", bty = "n", "PC 1")
```

How many PCs would you keep? Let's try 200 (from a possible 1000) and recreate the first face in our dataset.


```{r recon_four,  fig.width = 10,fig.height = 5}
n_pc <- 200
recon <- rowSums(pca$x[,1:n_pc]%*%pca$rotation[1,1:n_pc])
## add the "mean face" to reconstruct the original
par(mfrow = c(1,2), mar = c(0,0,0,0), oma = c(0,0,0,0))
## original
plot(face_set[[1]])
legend("bottom", bty = "n", paste(names(face_set[1]), "original", sep = " "))
## reconstructed
plot(pixmapGrey(recon + mean, 64, 64, bbox = c(0, 0, 64, 64)))
legend("bottom", bty = "n", paste(names(face_set[1]), "reconstructed", sep = " "))
```